---
title: "Project Report"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Background

New York City is infamous for the rats that live among us. Our apartments, buildings, subway stations, trash cans, parks, and streets are home to millions of rats and other [rodents](https://www.nytimes.com/2021/11/05/nyregion/nyc-rats-sightings.html). For many New Yorkers, rats are an inevitable part of everyday life, and one that has worsened in recent years. This year alone (2022), New York's Sanitation Department has reported more than 21,600 rat complaints. This represents a 71% increase since [October 2020](https://www.npr.org/2022/10/20/1130150756/new-york-rats-trash-rule-mayor-eric-adams). The issue has gotten so bad that Mayor Adams is now looking to hire a [“bloodthirsty” director](https://www.msn.com/en-ca/news/world/nyc-is-looking-for-bloodthirsty-rat-czar-and-the-job-pays-228-000/ar-AA14Qidf?li=AAggFp4) of rodent mitigation, who is eligible to make up to $228,000. Given the prevalence of rats in this city and the recent urgency of the issue, this project primarily seeks to explore trends in rat sightings across New York City.

Further, [research](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2862381/) shows that exposure to rodents is associated with various diseases and poor health outcomes, including asthma symptoms. Asthma symptoms have been increasing in middle to high-income countries, making this issue ever more critical . Understanding it can aid public health and health care combat diseases from these exposures. This is particularly an issue among communities that are historically and continually oppressed through racism and classism as they live in environments with higher levels of exposure due to air pollution, old buildings, cigarette smoke, and other factors. This project secondarily seeks to explore associations between rodent infestations and asthma prevalence in New York City.

A 2019 [study](https://towardsdatascience.com/rat-city-visualizing-new-york-citys-rat-problem-f7aabd6900b2) examined what neighborhoods are “rat hotspots.” It focused on data from 2010 to 2017. It looked at changes in number of sightings over the years as well as sightings by neighborhood. In 2016 there was a peak in the number of sightings. Brooklyn seemed to have the most sightings compared to other NYC boroughs. And they mostly occurred in 3+ family apartment buildings. The study also looked at high density areas - for example parts of the Upper West Side had a higher density of sightings compared to the Upper East Side. We will explore similar topics, using current data for 2022.


# Questions

This project is motivated by the following questions: 

* What trends are there in rat sightings throughout NYC?
* How do rat sightings vary by borough?
* How do rat sightings vary by location type?
* How are rodent trends associated with asthma trends?

# Data and Cleaning Process

This project uses two main datasets for our analysis.

**Rat Sightings**

The [Rat Sightings data](https://data.cityofnewyork.us/Social-Services/Rat-Sightings/3q43-55fe) is from NYC OpenData. This dataset includes information for all rat-related 311 service requests from 2010 to the present. We stored the data [locally? G drive?] since it was too large to upload to github. To clean the data, we changed the created_date variable from MM/DD/YYYY HH:MM into separate “month,” “day,” and “year” variables; made the year variable numeric; corrected capitalization fro address_type, city, and borough; recoded values for clarity; and filtered to remove observations in unspecified boroughs.  After cleaning the data, our key variables are: 

* month: month of rat sighting service request
* day: day of rat sighting service request
* year: year of rat sighting service request
* location_type: location type of rat sighting
* address_type: address type of rat sighting
* city: city of rat sighting
* borough: borough of rat sighting
* latitude: geo based latitude of incident location
* longitude: geo based longitude of incident location

**Asthma**

The [Asthma Data](https://webbi1.health.ny.gov/SASStoredProcess/guest?_program=/EBI/PHIG/apps/asthma_dashboard/ad_dashboard&p=it&ind_id=ad16) is from the New York State Asthma Dashboard. The asthma data set includes information on asthma rates in NYC. To clean the data, we renamed variables for clarity, updated variable types, and grouped information by year and borough. After cleaning, key variables include:

* year: year
* geography: borough
* number: number of asthma-related ER visits
* age_adjusted_rate_per_10_000: age-adjusted asthma-related ER visit rates per 10,000
* estimated_annual_rate_per_10_000: estimated asthma-related annual ER visit rate per 10,000

For our statistical analyses, we also merged the Rat Sightings and Asthma datasets by year and borough. 

# Exploratory Analysis

To better understand trends in rat sightings in NYC, we conducted some exploratory analysis using data visualization. 

First, we plotted rat sightings over time. Next, we explored the distribution of rat sighting frequency by borough and over time. Then, we explored top location types of rat sightings (i.e. apartment versus street versus park). Finally, we mapped rat sightings across the city using a density map. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(plotly)
library(readr)
library(patchwork)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%")

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis")

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r rat data, include=FALSE}
rat_df = read_csv("Rat_Sightings.csv") %>%
  janitor::clean_names() %>%
  mutate(created_date = gsub(" .*","", created_date),
         address_type = str_to_title(address_type),
         city = str_to_title(city),
         borough = str_to_title(borough),
         location_type = recode(location_type, "Other (Explain Below)" = "Other")) %>%
  separate(created_date, into = c("month", "day", "year"), sep = "/") %>%
  mutate(year = as.numeric(year),
         month = as.numeric(month),
         day = as.numeric(day)) %>%
  filter(borough != "Unspecified") %>%
  select(month, day, year, location_type, address_type, city, borough, latitude, longitude)
```

```{r asthma data, include=FALSE}
asthma_df = read_csv("asthma_ER_visits_adults.csv") %>%
  janitor::clean_names() %>%
  rename(
    year = time
  ) %>%
   mutate(age_adjusted_rate_per_10_000 = as.numeric(age_adjusted_rate_per_10_000),
         estimated_annual_rate_per_10_000 = as.numeric(estimated_annual_rate_per_10_000),
         number = as.numeric(number)) %>%
  filter(geo_type == "Borough") %>%
  group_by(year, geography) %>%
  select(year, geography, age_adjusted_rate_per_10_000, estimated_annual_rate_per_10_000, number)
```


```{r merge, include=FALSE}
asthma_rat_df <- left_join(rat_df, asthma_df, by = c('year' = 'year', 'borough' = 'geography'))
```

## Sightings over time

### New York City

First, we wanted to get a sense of rat sighting trends over time. The following line graph shows how the number of sightings have changed over time:

```{r, echo=FALSE}
overall_rat_line =
  rat_df %>%
  group_by(year) %>%
  count() %>%
  summarise(n_obs = n) %>% 
  ggplot(aes(x = year, y = n_obs)) + 
  scale_x_continuous(breaks = seq(2010, 2022, 2)) +
  geom_line() +
  labs(
    title = "Rat Sightings Over Time in NYC",
    x = "Year",
    y = "Number of Sightings")

overall_rat_line
```

Based on the plot, we see that rat sightings have generally increased since 2010, except for a brief decline between 2017 and 2020. Since 2020, the number of rat sightings in NYC has sharply increased. 

### By Borough

With a better understanding of general trends, we wanted to explore rat sighting trends over time by borough. The following line graph shows how the number of sightings has changed over time, color-coded by borough:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
rat_line = rat_df %>% 
  group_by(borough, year) %>%  
  count() %>%
  summarise(n_obs = n) %>% 
  ggplot(aes(x = year, y = n_obs , color = borough )) + 
  geom_line() +
  scale_x_continuous(breaks = seq(2010, 2022, by = 1)) +
  labs(
    title = "Rat Sightings Over Time by Borough",
    x = "Year",
    y = "Number of Sightings")

rat_line
```

Based on the plot, we see that the Bronx, Brooklyn, Manhattan, and Queens generally follow similar trends as NYC overall. Of these boroughs, the Bronx deviates the most, showing a decrease in rat sightings since 2021. Staten Island is different from all other boroughs and NYC overall. Rat sightings there have remained steady, and low, since 2010.


## Sightings Frequency

### By Borough

The following bar graph shows the distribution of rat sighting frequency by borough between 2010 and 2022:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
rat_bar = 
  rat_df %>% 
  count(borough) %>% 
  mutate(borough = fct_reorder(borough, n)) %>% 
  ggplot(aes(x = borough, y = n, fill = borough)) + 
  geom_bar(stat = "identity") +
  labs(
    title = "Frequency of Rat Sightings by Borough (2010-2022)",
    x = "Borough",
    y = "Number of Sightings",
    fill = "Borough")
  
  #plot_ly(x = ~borough, y = ~n, color = ~borough, type = "bar", colors = "viridis")

rat_bar
```

The plot shows that Brooklyn has the most rat sightings overall, with more than 70,000 sightings reported. Staten Island has the least number of sightings, with less than 10,000 sightings reported.

### By Borough and Year

This bar graph shows the distribution of rat sighting frequency by borough for each year with available data.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
rat_bar_time =
  rat_df %>% 
  group_by(borough) %>%
  count(year) %>% 
  ggplot(aes(x = year, y = n, fill = borough)) +
  geom_bar(stat = "identity",
           position = "dodge") +
  scale_x_continuous(breaks = seq(2010, 2022, by = 1)) +
  labs(
    title = "Frequency of Sightings in Boroughs Over Time",
    x = "Year",
    y = "Number of Sightings",
    fill = "Borough")

rat_bar_time
```

Since 2010, the overall frequency ranking of each borough has remained the same: from least to greatest, Staten Island, Queens, The Bronx, Manhattan, and Brooklyn. Since 2020, sightings have increased most significantly in Manhattan and Brooklyn. With the Bronx’s decrease since 2021, the Bronx and Queens are nearly tied so far in 2022.


## Top Location Types

### New York City

The bar graph shows the distribution of rat sightings frequency by location type. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
location_bar =
  rat_df %>%
  count(location_type) %>%
   mutate(
    location_type = fct_reorder(location_type, n),
    ranking = min_rank(desc(n))) %>% 
  filter(ranking <= 10) %>% 
  arrange(n) %>%
  ggplot(aes(x = location_type, y = n, fill = location_type)) + 
  geom_bar(stat = "identity") +
  labs(title = "Top 10 Location Types",
       x = "Location Type",
       y = "Number of Sightings",
       fill = "location_type") + coord_flip() +
  theme(legend.position = "none") 

location_bar
```

Overall, 3+ family apartment buildings had the highest number of sightings followed by 1-2 family dwellings. Of the top 10 location types, parking lots/garages had the lowest. 

### By Borough

The following bar graph shows the distribution of number of sightings by top sighting locations by borough:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
location_grid =
  rat_df %>%
  group_by(borough, location_type) %>%
  count() %>%
  summarise(n_obs = n) %>%
   mutate(
    location_type = fct_reorder(location_type, n_obs),
    ranking = min_rank(desc(n_obs))) %>% 
  filter(ranking <= 3) %>% 
  arrange(ranking) %>%
  ggplot(aes(x = location_type, y = n_obs, fill = location_type)) + 
  geom_bar(stat = "identity") +
  facet_grid(. ~ borough) +
  labs(title = "Top Sighting Locations by Borough",
       x = "Location Type",
       y = "Number of Sightings",
       fill = "location_type") +
  theme(axis.text.x = element_blank())
  
  
  #theme(axis.text.x = element_text(angle = 90, hjust = 1))

location_grid
```

In all boroughs except Queens and Staten Island, the highest number of sightings were in 3+ family apartment buildings. In Queens and Staten Island, the highest number of sightings were in 1-2 family dwellings.  

## Mapping

Finally, we mapped rat sightings across NYC by density: 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
rat_df %>%
  plot_ly(
    type = 'densitymapbox',
    lat = ~latitude,
    lon = ~longitude,
    coloraxis = 'coloraxis',
    radius = 10,
    color = ~borough) %>%
  layout(
    mapbox = list(
      style = "stamen-terrain",
      zoom = 9,
      center = list(lon = -73.9, lat = 40.7)), 
    coloraxis = list(colorscale = "Viridis"), 
    title = ("Density Map of Rat Sightings"))
```

As expected, we see rat sightings all over the city, with more rat sightings occurring in higher-density areas in the high-sighting boroughs identified in our previous plots. 

## Asthma trends

Next, we wanted to get a sense of asthma trends in New York City.

### Total Number of Asthma ER Visits Per Year

The line graph shows the total number of ER visits related to asthma per year in NYC. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
asthma_num_df =
  asthma_df %>%
  mutate(
    number = as.numeric(number), 
    data = "Total asthma ER visits"
  ) %>%
  group_by(year, data) %>%
  summarise(n_obs = sum(number, na.rm = TRUE))

asthma_rate_df =
  asthma_df %>%
  mutate(
    age_adjusted_rate_per_10_000 = as.numeric(age_adjusted_rate_per_10_000),
    data = "Age adjusted asthma ER visit rate"
  ) %>%
  group_by(year, data) %>%
  summarise(n_obs = mean(age_adjusted_rate_per_10_000, na.rm = TRUE))

asthma_estimated_df =
  asthma_df %>%
  mutate(
    estimated_annual_rate_per_10_000 = as.numeric(estimated_annual_rate_per_10_000), 
    data = "Estimated annual asthma rate"
  ) %>%
  group_by(year, data) %>%
  summarise(n_obs = mean(estimated_annual_rate_per_10_000, na.rm = TRUE))

asthma_years_num_plot =
  asthma_num_df %>%
  ggplot(aes(x = year, y = n_obs)) + 
  geom_line() +
  scale_x_continuous(breaks = seq(2005, 2018, 1)) +
  labs(title = "Total number of asthma ER visits per year",
       x = "Year",
       y = "Total number of asthma ER visits")
  

asthma_rate_estimated_plot = 
  ggplot(data = asthma_rate_df, aes(x = year, y = n_obs, color = data)) + 
  geom_line() +
  geom_line(data = asthma_estimated_df) +
  scale_x_continuous(breaks = seq(2005, 2018, 1)) +
  labs(title = "Age adjusted vs. Estimated Asthma Rates Over The Years",
       x = "Year",
       y = "Rate per 10,000 people")
  

asthma_years_num_plot
```

Overall the total number of asthma-related ER visits rises steadily. We see 0 visits in 2015 because the dataset had no data for that year. 

### Age-adjusted vs. estimated Asthma rates over time

The line graph compares age-adjusted ER asthma rates over time to the estimated ER asthma rate. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
asthma_rate_estimated_plot
```

This allowed us to see that there was little difference between the two variables. 

###  Mean Asthma rate by borough

The bar graph shows the mean ER asthma rates by borough through 2018. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
asthma_rate_borough_plot = 
  asthma_df %>% 
  group_by(geography) %>% 
  mutate(
    age_adjusted_rate_per_10_000 = as.numeric(age_adjusted_rate_per_10_000),
    geography = as.factor(geography)
  ) %>%
  summarise(
    asthma_rate = mean(age_adjusted_rate_per_10_000, na.rm = TRUE)
  ) %>%
  mutate(geography = fct_reorder(geography, asthma_rate)) %>% 
  ggplot(aes(x = geography, y = asthma_rate, fill = geography)) + 
  geom_bar(stat = "identity") +
  labs(
    title = "Mean asthma rate per borough (2004-2018)",
    x = "Borough",
    y = "Mean Age-Adjusted Asthma Rate per 10,000",
    fill = "Borough")

asthma_rate_borough_plot
```

The Bronx has the highest asthma rate, followed by Manhattan and Brooklyn. Queens has the lowest rate. 

### Mean Asthma rate by the borough over time

The bar graph shows the mean ER asthma rates by borough over time.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
asthma_borough_time_plot = 
  asthma_df %>% 
  group_by(geography, year) %>% 
  mutate(
    age_adjusted_rate_per_10_000 = as.numeric(age_adjusted_rate_per_10_000),
    geography = as.factor(geography)
  ) %>%
  summarise(
    asthma_rate = mean(age_adjusted_rate_per_10_000, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = year, y = asthma_rate, fill = geography)) +
  geom_bar(stat = "identity",
           position = "dodge") +
  scale_x_continuous(breaks = seq(2005, 2018, by = 1)) +
  labs(
    title = "Mean Asthma ER Visits in Boroughs Over Time",
    x = "Year",
    y = "Mean Asthma ER Visits per 10,000",
    fill = "Borough")

asthma_borough_time_plot
```

The Bronx has the highest asthma rate, followed by Manhattan and Brooklyn. Queens has the lowest rate. The distribution of rates generally remains the same as what we saw in the overall mean asthma rate by borough. However, over time, we see that Brooklyn surpasses Manhattan’s asthma rate. 

## Asthma versus Rat sighting

Finally, we wanted to start looking at associations between the two datasets. The side-by-side line plots show the mean age-adjusted ER visits rate per 10,000 of asthma over time next to rat sightings over time. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
asthma_line =
  asthma_rat_df %>%
  filter(year > 2009) %>%
  mutate(age_adjusted_rate_per_10_000 = as.numeric(age_adjusted_rate_per_10_000, na.rm = TRUE)) %>%
  group_by(year) %>%
  summarize(mean_age_adjust = mean(age_adjusted_rate_per_10_000)) %>%
  drop_na(mean_age_adjust) %>%
  ggplot(aes(x = year, y = mean_age_adjust)) + 
  geom_line() +
  labs(
    title = "Asthma ED Visits Over Time in NYC",
    x = "Year",
    y = "Mean age adjusted rate per 10,000")

asthma_line + overall_rat_line
```



# Statistical Analysis


# Discussion
